---
title: "Laplace transform and soils"
author: "Chris H Wilson"
date: "10/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# Background 

Soil organic carbon continas a heterogeneous assortment of organic molecules interacting with the soil mineral matrix in a variety of ways. This heterogeneity has traditionally been represented in process-based models (PBMs) using multiple discrete pools, characterized either by inherently varying turnover times, perhaps as a function of environmental variables like temperature and soil moisture (e.g. CENTURY, Parton 1988), or by differing mechanisms of protection (e.g. MIMICS, Wieder 2014). By contrast, a recent paper (Waring et al. 2020) has proposed an individual-based model (IBM) framework in which cohorts of carbon atoms are tracked probabilistically over time. Regardless of approach, a key insight emphasize by Waring et al. (2020), but latent within other approaches (e.g. Wieder 2014), is that carbon residence times are best thought of as *emergent properties* rather than *inherent properties*, whether of different classes of inputs or discrete "pools" of SOC. 

Here, we propose to take this *emergent property* view one step further. From an applied point of view, one pressing goal is to quantify the carbon sequestration service accomplished by various alternative management practices that aim to increase or maintain SOC. For this outcome, we propose the ideal PBM output is a *distribution over residence times*, such as Fig. X in Waring et al. (2020). We show in this short note how the residence time distribution can be upscaled to a service quantification via an integral transform. The virtue of this approach is that PBM modeling can investigate the impact of a variety of model formulations, stabilization processes, and stochasticity assumptions, while maintaining a **clear goal**: estimate the distribution of residence times characterizing either a heterogeneous pool or a cohort of carbon atoms. In addition to clarity of purpose, our mathematics will drastically simplify the computational challenges of, say, attempting to fit an IBM over a large landscape, or globally. Indeed, the focus for computation can be reduced to adequately representing a distribution over residence times. 

# Integral transform

Given a distribution ($\pi$) of residence times ($\lambda$) $\pi(\lambda)$ for a given class of organic molecules within a soil, the contribution of a cohort of carbon atoms $C$ to offseting atmospheric carbon is simply the fraction remaining in the soil at time $t$. This can be represented as an expectation: $$\tag{1} \int{C\pi(r)e^{-\frac{1}{\lambda}t}dr}$$ which turns out simply to be the Laplace transform of the random variable $r$. Given the generally recognized right-skewed distribution, an inverse-gamma distribution is a natural choice. Thus, the integral transform is:
<!-- $$\tag{2} \int{C\frac{\beta^\alpha}{\Gamma(\alpha)}r^{\alpha-1}e^{-\beta r}e^{-rt}}dr$$ --> 
$$\tag{2_b} \int{C\frac{\beta^\alpha}{\Gamma(\alpha)}\lambda^{-\alpha-1}e^{-\frac{\beta}{\lambda}}e^{-\frac{1}{\lambda}t}}dr$$

Fortunately, this integral has a very simple analytical solution, dropping the cohort constant "C":
$$\tag{3} (\frac{\beta}{\beta+t})^\alpha$$
We reparameterize this solution because, while simple, the parameters of the inverse-gamma distribution $\beta$ and $alpha$ do not necessarily have easy intuition. However, we can translate into a parameterization with a mean $\mu$ and a dispersion constant $k$, and rewrite the integral transform in (3). To do this, we first note that the first moment (mean) of an inverse gamma is given by $\frac{\beta}{\alpha-1}$, while the second moment (variance) is $\frac{\beta^2}{(\alpha-1)^2(\alpha-2)}$. The first moment is only defined when $alpha$>1, and the second moment only when $alpha$>2. Thus, for our reparameterization, we will set the restriction that $alpha$ is strictly > 2. T







```{r}

mu <- 4
k <- 10

resFunc <- function(mu=mu,k=k,t){
  invK <- 1/k
  left <- (invK/(invK + t))^(mu*k); 
}


resFunc2 <- function(mu=mu,k=k,t){
  invK <- 1/k
  left <- ((mu*(invK+1))/((mu*(invK+1))+t))^(invK+2); 
}

ggplot(data=data.frame(t=c(0,20)),aes(x=t)) + 
  stat_function(fun=resFunc2,args=list(mu=mu,k=k)) + 
  stat_function(fun=resFunc2,args=list(mu=mu,k=k*0.1),color="blue")





```

