---
title: "Laplace transform and soils"
author: "Chris H Wilson"
date: "10/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# Background 

Soil organic carbon consists of a heterogeneous assortment of organic molecules phyisically protected from microbial decomposition through soil mineral interactions. This heterogeneity has traditionally been represented in process-based models (PBMs) using multiple discrete pools differentiated on the basis of residence time, e.g. the labile, slow and passive pools in CENTURY (Parton 1988), or by differing mechanisms of protection, e.g. the physical versus biochemical pools in MIMICS (Wieder 2014). Both approaches are zero order approximations of a necessarily far richer reality. For this reason, efforts to operationalize these various "pools" in the form of empirical procedures (e.g. size-density fractionation) have not generally been satisfying. Stable isotope probing has suggested that any such operationally defined "pool" is itself heterogeneous (citations: our work at BRU, others). Recently, Waring et al. (2020) propose an individual-based model (IBM) framework in which cohorts of carbon atoms are tracked probabilistically over time, as they transform from one biochemical state to another, and participate in various soil mineral interactions. One happy feature of this approach is that residence time is correctly represented as an  *emergent property*, rather than identified as an inherent feature of a supposedly homogeneous pool. In fact, the output of the Waring et al. (2020) model is a *distribution over residence times*, for each major class of organic molecule, thus providing one promising path to embracing real-world SOC heterogeneity. However, regardless of **where** the heterogeneity comes from, or **how** it is represented, there is a clear need for a consistent mathematical framework that will allow us to deduce the consequences of this heterogeneity for long-term SOC dynamics, and prediction of the ecosystem services that depend on SOC, such as greenhouse gas mitigation. Below, I outline a generic, quantitative framework that maps a distribution over SOC residence times to a scaled-up dynamics via use of an integral transform. I derive two universal curves bounding the possible scaled-up dynamics, and provide a clear path to map from fitted models to ecosystem service. 


# Integral transform of heterogeneous SOC 

Given a distribution ($\pi$) of residence times ($\lambda$) $\pi(\lambda)$ for a given class of organic molecules within a soil, the contribution of a cohort of carbon atoms $C$ to offseting atmospheric carbon is simply the fraction remaining in the soil at time $t$. This can be represented as an expectation: $$\tag{1} \int{C\pi(r)e^{-\frac{1}{\lambda}t}dr}$$ which turns out simply to be the Laplace transform of the random variable $r$. Given the generally recognized right-skewed distribution, an inverse-gamma distribution is a natural choice. Thus, the integral transform is:
<!-- $$\tag{2} \int{C\frac{\beta^\alpha}{\Gamma(\alpha)}r^{\alpha-1}e^{-\beta r}e^{-rt}}dr$$ --> 
$$\tag{2_b} \int{C\frac{\beta^\alpha}{\Gamma(\alpha)}\lambda^{-\alpha-1}e^{-\frac{\beta}{\lambda}}e^{-\frac{1}{\lambda}t}}dr$$

Fortunately, this integral has a very simple analytical solution, dropping the cohort constant "C":
$$\tag{3} (\frac{\beta}{\beta+t})^\alpha$$
We reparameterize this solution because, while simple, the parameters of the inverse-gamma distribution $\beta$ and $alpha$ do not necessarily have easy intuition. However, we can translate into a parameterization with a mean $\mu$ and a dispersion constant $k$, and rewrite the integral transform in (3). To do this, we first note that the first moment (mean) of an inverse gamma is given by $\frac{\beta}{\alpha-1}$, while the second moment (variance) is $\frac{\beta^2}{(\alpha-1)^2(\alpha-2)}$. The first moment is only defined when $alpha$>1, and the second moment only when $alpha$>2. Thus, for our reparameterization, we will set the restriction that $alpha$ is strictly > 2. The first moment 







```{r}

mu <- 4
k <- 10

resFunc <- function(mu=mu,k=k,t){
  invK <- 1/k
  left <- (invK/(invK + t))^(mu*k); 
}


resFunc2 <- function(mu=mu,k=k,t){
  invK <- 1/k
  left <- ((mu*(invK+1))/((mu*(invK+1))+t))^(invK+2); 
}

ggplot(data=data.frame(t=c(0,20)),aes(x=t)) + 
  stat_function(fun=resFunc2,args=list(mu=mu,k=k)) + 
  stat_function(fun=resFunc2,args=list(mu=mu,k=k*0.1),color="blue")





```

