---
title: "Universal Curves to Represent Heterogeneity in Soil Carbon Residence Times"
author: "Chris H Wilson"
date: "10/18/2020"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# Background 

Soil organic carbon consists of a heterogeneous assortment of organic molecules phyisically protected from microbial decomposition through soil mineral interactions. This heterogeneity has traditionally been represented in process-based models (PBMs) using multiple discrete pools differentiated on the basis of residence time, e.g. the labile, slow and passive pools in CENTURY (Parton 1988), or by differing mechanisms of protection, e.g. the physical versus biochemical pools in MIMICS (Wieder 2014). Both approaches are zero order approximations of a necessarily far richer reality. For this reason, efforts to operationalize these various "pools" in the form of empirical procedures (e.g. size-density fractionation) have not generally been satisfying. Stable isotope probing has suggested that any such operationally defined "pool" is itself heterogeneous (citations: our work at BRU, others). Recently, Waring et al. (2020) propose an individual-based model (IBM) framework in which cohorts of carbon atoms are tracked probabilistically over time, as they transform from one biochemical state to another, and participate in various soil mineral interactions. One happy feature of this approach is that residence time is correctly represented as an  *emergent property*, rather than identified as an inherent feature of a supposedly homogeneous pool. In fact, the output of the Waring et al. (2020) model is a *distribution over residence times*, for each major class of organic molecule, thus providing one promising path to embracing real-world SOC heterogeneity. However, regardless of **where** the heterogeneity comes from, or **how** it is represented, there is a clear need for a consistent mathematical framework that will allow us to deduce the consequences of this heterogeneity for long-term SOC dynamics, and prediction of the ecosystem services that depend on SOC, such as greenhouse gas mitigation. Below, I outline a generic, quantitative framework that maps a distribution over SOC residence times to a scaled-up dynamics via use of an integral transform. I derive two universal curves bounding the possible scaled-up dynamics, and provide a clear path to map from fitted models to ecosystem service. 


# Scaling up over SOC heterogeneity: integral transform 

Given a distribution ($\pi$) of residence times ($\lambda$) $\pi(\lambda)$ for a given class of organic molecules within a soil, the contribution of a cohort of carbon atoms $C$ to offseting atmospheric carbon is simply the fraction remaining in the soil at time $t$. Rather than rely on expensive computational experiments, this can be solved analytically, with tremendous payoff. We represent the scaled-up dynamics as an expectation over the heterogeneity: $$\tag{1} \int{C\pi(r)e^{-\frac{1}{\lambda}t}dr}$$ which turns out to be essentially the Laplace transform of the random variable $\lambda$. 

In short, the framework proposed here invokes two assumptions. First, our cohort of carbon atoms is sufficiently large that we can apply a "mass action" assumption. That is, we represent their possible fates by the residence time distribution $\pi(\lambda)$ convoluted with turnover via the exponential in equation (1). Basically, this is a simple first order decay assumption at the level of the individual, while *the behavior of the ensemble* requires the integral transform in (1). Secondly, we start with the observation that the heterogeneity in SOC residence times (Waring et al. 2020) is generally right-skewed and unimodal, and we capture these features via an inverse-gamma distribution with mild restrictions. 

With these assumptions in place, the specific integral transform is:
<!-- $$\tag{2} \int{C\frac{\beta^\alpha}{\Gamma(\alpha)}r^{\alpha-1}e^{-\beta r}e^{-rt}}dr$$ --> 
$$\tag{2_b} \int{C\frac{\beta^\alpha}{\Gamma(\alpha)}\lambda^{-\alpha-1}e^{-\frac{\beta}{\lambda}}e^{-\frac{1}{\lambda}t}}dr$$

Fortunately, this integral has a very simple analytical solution, dropping the cohort constant "C":
$$\tag{3} (\frac{\beta}{\beta+t})^\alpha$$
We reparameterize this solution because, while simple, the parameters of the inverse-gamma distribution $\beta$ and $alpha$ do not necessarily have easy intuition. However, we can translate into a parameterization with a mean $\mu$ and a dispersion constant $k$, and rewrite the integral transform in (3). To do this, we first note that the first moment (mean) of an inverse gamma is given by $\frac{\beta}{\alpha-1}$, while the second moment (variance) is $\frac{\beta^2}{(\alpha-1)^2(\alpha-2)}$. The first moment is only defined when $alpha$>1, and the second moment only when $alpha$>2. Thus, for our reparameterization, we will set the restriction that $alpha$ is strictly > 2. We notate the first moment as $\mu$, and retain this parameter. For second moment, note that: 
$$\tag{4} Var(\lambda) = \mu(\lambda)^2\frac{1}{\alpha-2}$$
So, we define a new parameter $\phi = \frac{1}{\alpha-2}$. For interpretation, $\phi$ turns out to be equivalent to the coefficient of variation (CV) squared, and thus we consider $phi$ a variability or dispersion constant. Solving for the parameters and substituting, the integral transform in (3) is now: 
$$\tag{5} \frac{\mu(\frac{1}{\phi}+1)}{\mu(\frac{1}{\phi}+1)+t}^{(\frac{1}{\phi}+2)}$$

To be extremely clear, the amount of carbon from a given cohort of carbon atoms remaining at time $t$ is given by (5): 
$$\tag{6} C(t) = C(0)\frac{\mu(\frac{1}{\phi}+1)}{\mu(\frac{1}{\phi}+1)+t}^{(\frac{1}{\phi}+2)}$$
Thus, given a time series of observations of turnover (e.g. from incubation experiments such as XXXX), we could estimate the relevant parameters governing the distribution of residence times $\mu$ and $\phi$ inversely by fitting to equation (6). Alternately, we can use forward simulation results from a suitable PBM that provides such a distribution (e.g. Waring et al. 2020), plug in the resulting parameters and use (6) to estimate scaled-up properties and services. 

# Universal bounding curves 

The function in (6) can be further probed for theoretical insight by non-dimensionalization and examination of limiting behavior. First, we consider the case where variablity becomes maximal, by taking the limit $\phi \to \inf$, resulting in: 
$$\tag{7} C(t) = C(0) \frac{\mu}{\mu+t}^2$$
We non-dimensionalize 7 with the quantity $\tau = \frac{t}{\mu}$, where $\tau$ is thus a dimensionless time representing multiples of the mean of the residence time distribution, and we have: 
$$\tag{8} C(t) = C(0) \frac{1}{1+\tau}^2$$





```{r}

mu <- 4
k <- 10


resFunc2 <- function(mu=mu,k=k,t){
  invK <- 1/k
  left <- ((mu*(invK+1))/((mu*(invK+1))+t))^(invK+2); 
}

ggplot(data=data.frame(t=c(0,20)),aes(x=t)) + 
  stat_function(fun=resFunc2,args=list(mu=mu,k=k)) + 
  stat_function(fun=resFunc2,args=list(mu=mu,k=k*0.1),color="blue")





```

